### 2013-01-24 patchy-selection-paper.tex
/* Patchy Selection paper */

\paragraph{Previous work} We will make use of a number previous
results from the literature. \citet{slatkin1973geneflow}, working in
one dimensional deterministic framework, showed that if the
physical width of the patch is less than $2 \tan^{-1}
(\sqrt{|s_m/s_B|})$, then $B$ can not 
become stably established within the patch, as the local adaptation prevented by
migrational swamping \citep[see also][ for a review]{Lenormand}.
 \citet{barton1987establishment}, again working in one dimension,
 showed that this critical patch size also held in a stochastic
 framework \citep[see also the work of][]{Polk}. 
For patches above this size, \citet{barton1987establishment}
 founds that mutations appearing further than a few multiples of
 $\sigma/\sqrt{2|s_m|}$ away from such a patch have vanishingly small
 probability of becoming established within the patch. 
Barton also showed that mutations appearing within the patch have probability 
strictly less but approaching that $2s_B$ of establishment as the size
of the patch increased. Thus for large enough patches the probability
of establishment of a locally favorable allele approaches the standard
establishment probability ($2s_B$) for a panamitic population
\citep[][]{haldane, fisher}, a result that also (generally) holds for
a geographically spread population experiencing a uniform selection
pressure \citep{Maruyama, cherry}.

@@

\subsubsection{Time to the establishment of a locally adaptive allele
  due to mutational influx}
\label{ss:patchymutation}
Consider first a single, isolated poor habitat patch of area $A$ in
which no $B$ allele has yet become established. As
we are interested in local adaptation that can occur we will assume
that our patch is considerably larger than the cutoff for local
establishment, i.e. $\sqrt{A} \gg 2 \tan^{-1} (\sqrt{|s_m/s_B|})$.

@@

The rate at which mutations arise and colonize the patch is
\begin{equation}
\lambda_{Mut} = \rho \mu p_f \approx 2 s_B \rho \mu.
\end{equation}

If this rate is low,  then the waiting time till a mutation arises that
will become locally established within the patch is exponentially
distributed with rate $\lambda_{Mut}$.  Assuming that once a mutation
becomes established
 it quickly reaches its equilibrium frequency across the patch, the
 expected time it takes for our patch to become colonized by the $B$
 allele is $T_{Mut} \approx 1/ \lambda_{Mut} $.

\subsubsection{Time to the establishment of a locally adaptive allele
  due to migrational influx}
\label{ss:patchymigration}
Now suppose that there are two patches, each of area $A$, of the poor
habitat type separated by distance $R$. Consider the case where the
$B$ allele has arisen and become established in the first patch, but has not yet appeared in the second.
As above, any copy of the new allele that appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.
Due to mutation--selection balance there will be a small number of $B$ alleles descended from those of the first patch (``migrants'') 
present in the regions where they are disadvantageous;
the equilibrium frequency may again be expressed in terms of Jacobi elliptic functions (Appendix XXX, citeXXX),
although the equations are again not particularly helpful.
However, we do know that if $q(r)$ is the expected proportion present at \gc{migration}--selection balance at distance $r$ from the first patch
in the absence of the second patch, that for large $r$ the proportion decays exponentially, $q(r) \approx C \exp( -|\sqrt{s_m} r / \sigma|)$,
where $C$ is a constant depending on the structure of the populations
and the selection coefficients. \gc{comment on size of C? Should we
  also use this approx. to get the approx. rate of establishing mutations
  arising outside of the patch, above. Feel like in the mutation rate
  part we could pull $s_B$ out front of all rates.}

@@

Combining this with the calculations above allows us to get a rough sense of the scale of migration between separated patches,
supposing that patches are large enough for local fixation to be stable. If the alleles were not advantageous in the second patch,
the average number of mutant alleles present in a small area $\Delta$
near the second patch, due
to migration from the first patch, would be around $\rho A \exp( -
|\sqrt{s_m} R/\sigma| ) \Delta$. Assuming that the patch is
sufficiently large to give an allele a $2s_B$ probability of becoming established, the rate at which migrants find
themselves in the patch and become established  is approximately
\begin{equation}
\lamda_{Mig,R} = \rho A^2 s_B \exp( -|\sqrt{s_m} R/\sigma| ).
\end{equation}
\gc{check my squared here. Do we want to use $q(R)$ here?|
If this establishment rate is low, the time it takes
a mutation to arise that will become established will be  $\sim Exp(\lambda_{mig,R})$

If  the time it takes for a mutation that it destined to 
become established in the second patch to become establish is short,
then  the expected waiting time until some migrant allele fixes in the
second patch will be $T_{Mig} = \lamda_{Mig,R} $.

@@

@@

\subsubsection{The probability of parallel adaptation between patches} 
\label{ss:probparallel}

@@

We will assume that once a introduced $B$ allele (from migration or mutation) becomes
established in the poor habitat patch it reaches its equilbrium
frequency very rapidly. Assuming that we being from our geographically 
spread population containing no copies of the $B$ allele,

Under this approximation, and our selective
exclusion assumption, the second patch will become colonized by migration or
mutation, but not by both. This means that the question of whether the
second patch adapt in parallel, or through a shared migrant allele,
simply comes down to whether the a new mutation or a migrant allele
becomes established first.

If environment is continuous, with separated patches where the new allele is advantageous,
we can only get a rough sense of whether the new adaptation is likely to spread through mutation (i.e.\ in parallel)
or through migration.
Taking the expressions from sections \ref{ss:patchymutation} and \ref{ss:patchymigration} above,
we see that if patches each have sufficiently large area $A$,
then the mutational input in each is on the order of $2 \rho \mu A$,
while the influx due to migration between patches separated by distance $R$ is on the order $2 \rho A \exp(- |\sqrt{s_m} R/\sigma|)$. 
If the input from mutation and migration are comparable, i.e.\ if $\mu \approx \exp( -|\sqrt{s_m} R/\sigma| )$,
then adaptation is expected to take place both by migration and
mutation.

If we take $\mu = 10^{-8}$, then this is satisfied if $R/\sigma \approx 18/\sqrt{s_m}$.
so if the disadvantage to the new mutation between patches is fairly strong (say, $s_m=.05$),
then the distance between patches must be about a hundred times the dispersal distance (i.e.\ $R/\sigma \approx 100$);
while if $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by many hundreds of times the dispersal distance (i.e.\ $R/\sigma \approx 500$).
The situation does not change much if $\mu$ is larger -- even if $\mu = 10^{-6}$, 
the required separation between patches is only reduced by 60\%.
If the separation between patches is much larger -- say, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will be independent between patches.
On the other hand, if the separation is much smaller, then a single mutation is likely to fix everywhere.

Furthermore, in the model of discrete demes connected by symmetric weak migration, 
the final distribution of types depends only on the ratio of mutation to migration rates,
and that the number of independently arisen types follows the Ewens distribution, with parameter $\mu/m$.
In Figure \ref{fig:discbyratio} we plot mean number of types and size (number of demes occupied) of a sampled type,
for a model with 20 interconnected demes, across different parameter values.  
This indicates that we need the population-scaled mutation and migration rates 
to be within a factor of about 100 of each other for parallel adaptation to leave an interesting pattern.  
If migration is faster, then a single type is likely to take over, while if migration is weaker, each deme is likely to come up with its own type.

>>>>

\paragraph{Previous work} We will make use of a number previous
results from the literature. \citet{slatkin1973geneflow}, working in
one dimensional deterministic framework, showed that if the
physical width of the patch is less than $2 \tan^{-1}
(\sqrt{|s_m/s_B|})$, then $B$ can not 
become \gc{fixed/at a stable equilibrium/?} within the patch, as the local adaptation prevented by
migrational swamping \citep[see also][ for a review]{Lenormand}.
 \citet{barton1987establishment}, again working in one dimension,
 showed that this critical patch size also held in a stochastic
 framework \citep[see also the work of][]{Polk}. 
For patches above this critical size, \citet{barton1987establishment}
 founds that mutations appearing further than a few multiples of
 $\sigma/\sqrt{2|s_m|}$ away from such a patch have vanishingly small
 probability of becoming established within the patch. 
Barton also showed that mutations appearing within the patch have probability 
strictly less but approaching that $2s_B$ of establishment as the size
of the patch increased. Thus for large enough patches the probability
of establishment of a locally favorable allele approaches the standard
establishment probability ($2s_B$) for a panamitic population
\citep[][]{haldane, fisher}, a result that also (generally) holds for
a geographically spread population experiencing a uniform selection
pressure \citep{Maruyama, cherry}.

@@

\subsubsection{Time to the establishment of a locally adaptive allele
  due to mutational influx}
\label{ss:patchymutation}
Consider first a single, isolated poor habitat patch of area $A$ in
which no $B$ allele has yet become established. As
we are interested in patchess where local adaptation can occur we will assume
that our patch is larger than the cutoff for local
establishment, i.e. $\sqrt{A} > 2 \tan^{-1} (\sqrt{|s_m/s_B|})$.

@@

The rate at which mutations arise and colonize the patch is
\begin{equation}
\lambda_{Mut} = \rho \mu p_f \approx 2 s_B \rho \mu.
\end{equation}
If this rate is low,  then the waiting time till a mutation arises that
will become locally established within the patch is exponentially
distributed with rate $\lambda_{Mut}$.  Assuming that once a mutation
becomes established it quickly reaches its equilibrium frequency across the patch, the
 expected time it takes for our patch to become colonized by the $B$
 allele is $T_{Mut} \approx 1/ \lambda_{Mut} $.

\subsubsection{Time to the establishment of a locally adaptive allele
  due to migrational influx}
\label{ss:patchymigration}
Now suppose that there are two patches, each of area $A$, of the poor
habitat type separated by distance $R$. Consider the case where the
$B$ allele has arisen and become established in the first patch, but has not yet appeared in the second.
As above, any copy of the new allele that appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.
Due to mutation--selection balance there will be a small number of $B$ alleles descended from those of the first patch (``migrants'') 
present in the regions where they are disadvantageous;
the equilibrium frequency may again be expressed in terms of Jacobi elliptic functions (Appendix XXX, citeXXX),
although the equations are again not particularly helpful.
However, we do know that if $q(r)$ is the expected proportion present at \gc{migration}--selection balance at distance $r$ from the first patch
in the absence of the second patch, that for large $r$ the proportion decays exponentially, $q(r) \approx C \exp( -|\sqrt{s_m} r / \sigma|)$,
where $C$ is a constant depending on the structure of the populations
and the selection coefficients. \gc{comment on size of C? Should we
  also use this approx. to get the approx. rate of establishing mutations
  arising outside of the patch, above. Feel like in the mutation rate
  part we could pull $s_B$ out front of the rate as well.}

@@

Combining this with the calculations above allows us to get a rough sense of the scale of migration between separated patches,
supposing that patches are large enough for local fixation to be stable. If the alleles were not advantageous in the second patch,
the average number of mutant alleles present in a small area $\Delta$
near the second patch, due
to migration from the first patch, would be $\approx \rho A \exp( -
|\sqrt{s_m} R/\sigma| ) \Delta$. Assuming that the patch is
sufficiently large to give an allele a $2s_B$ probability of becoming established, the rate at which migrants find
themselves in the patch and become established  is approximately
\begin{equation}
\lamda_{Mig,R} = \rho A^2 s_B \exp( -|\sqrt{s_m} R/\sigma| ).
\end{equation}
\gc{Is this $A^2$ right here?. Do we want to use $q(R)$ here?}
If this establishment rate is low, the time it takes
a mutation to arise that will become established will be  $\sim Exp(\lambda_{mig,R})$
Assuming that the time it takes for a allele from the first patch to
become established in the second patch is short,
then  the expected waiting time until some migrant allele fixes in the
second patch will be $T_{Mig} = 1/\lamda_{Mig,R} $.

@@

\gc{Put time it takes for a migrant allele to move between patches
  here? Or below }

@@

\subsubsection{The probability of parallel adaptation between patches.} 
\label{ss:probparallel}
We now turn to the question of whether our populations adapt in
parallel genetic changes or by the spread of a migrant allele between the
patches. These two scenerios are assumed to be distinguishable, either
because in the case of parallelism the mutations occur at different
positions or if the mutations occur at the same position they can be
distinguished due to their haplotypic background.

Assuming that we begin with our geographically 
spread population containing no copies of the $B$ allele, one or other
of our poor habitat patches will be the first to adapt via the
establishment of a $B$ allele. The question

@@

We will assume that once a introduced $B$ allele (from migration or mutation) becomes
established in the poor habitat patch it reaches its equilbrium
frequency very rapidly. Under this approximation, and our selective
exclusion assumption, after one of the patches becomes colonized by a
single $B$ allele, the other patch will become colonized by migration or
mutation, but not by both. As such, the question of whether the
population adapts to the second poor habitat patch in parallel, simply comes down to whether the a new mutation or a migrant allele
is the first to become established in the second patch.

Taking the expressions from sections \ref{ss:patchymutation} and \ref{ss:patchymigration} above,
we see that if patches each have sufficiently large area $A$,
then the mutational input in each is on the order of $2 s_B \rho \mu A$,
while the influx due to migration between patches separated by
distance $R$ is on the order $2 s_B \rho A \exp(- |\sqrt{s_m}
R/\sigma|)$ \gc{check $A^2$}. 
If the input from mutation and migration are comparable, i.e.\ if $\mu \approx \exp( -|\sqrt{s_m} R/\sigma| )$,
then adaptation is expected to take place both by migration and
mutation.

If we take $\mu = 10^{-8}$, then this is satisfied if $R/\sigma \approx 18/\sqrt{s_m}$.
so if the disadvantage to the new mutation between patches is fairly strong (say, $s_m=.05$),
then the distance between patches must be about a hundred times the dispersal distance (i.e.\ $R/\sigma \approx 100$);
while if $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by many hundreds of times the dispersal distance (i.e.\ $R/\sigma \approx 500$).
The situation does not change much if $\mu$ is larger -- even if $\mu = 10^{-6}$, 
the required separation between patches is only reduced by 60\%.
If the separation between patches is much larger -- say, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will be independent between patches.
On the other hand, if the separation is much smaller, then a single
mutation is likely to fix in both patches.

If we are willing to take these approximations at face
value, and assuming both rates are small, the probability of parallel
adaptation is 
\begin{equation}
\frac{\mu}{\mu + \exp(- |\sqrt{s_m} R/\sigma|}
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches at a
rate $\sqrt{s_m} /\sigma$. However, we caution that these calculations
are rough and so should be taken wi

%Furthermore, in the model of discrete demes connected by symmetric weak migration, 
%the final distribution of types depends only on the ratio of mutation to migration rates,
%and that the number of independently arisen types follows the Ewens distribution, with parameter $\mu/m$.
%In Figure \ref{fig:discbyratio} we plot mean number of types and size (number of demes occupied) of a sampled type,
%for a model with 20 interconnected demes, across different parameter values.  
%This indicates that we need the population-scaled mutation and migration rates 
%to be within a factor of about 100 of each other for parallel adaptation to leave an interesting pattern.  
%If migration is faster, then a single type is likely to take over, while if migration is weaker, each deme is likely to come up with its own type.
